A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-06-24 rigid_bodies.tex
first modified version Ana

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

\begin{document}

@@

Improvements in molecular dynamics algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations. This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

@@

one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient. In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble. The property of being symplectic implies the conservation of volume in phase space. This feature, together with time reversibility are mandatory for numerical integrators used in hybrid monte carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion. This approach has been successfully used in pure molecular dynamics as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}. It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.

In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform molecular dynamics. The contribution for the NVE case relies on the factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}. In the NVT case, both the translational and rotational degrees of freedom (DOF) of all particles are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF. It should be mentioned here that there are several numerical integrators\cite{Omelyan2007,Omelyan2008,vanZon2008} for the equations of motion for rigid bodies in the NVE ensemble which are more accurate than the symplectic Verlet-type integrators. However, since the application of that methods is limited to Hamiltonian systems, it's unclear whether their benefits will extend to the numerical integrators in the NVT ensemble. This, together with its great applicability, are the main reasons why our approach is based on the Verlet-type integrators.

We have also derived an expression for calculating the pressure of a rigid body system whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions. That expression depends on the resultant interaction forces between rigid body pairs. This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces. This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. As a preamble to the integration schemes, we show an alternative way of interpreting the action of the Liouville operators on the phase space variables. In Sec.~\ref{sec:moleculardynamics} we present the factorization of the Liouville operator for both the NVE and NVT ensembles. Following this, in Sec.~\ref{sec:pressure} we show the pressure expression developed here and we compare the result with that obtained from other method. In Sec.~\ref{sec:performance} we discuss the performance of the numerical integrators according to the conservation of the total energy. For the NVT ensemble, we also present the results from a test that checks if the potential and kinetic energies are actually sampled according to the canonical distribution. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation.

\section{Mechanics of Rigid Body Motion}
\label{sec:mechanics}
In this section we review one particular factorization of the rotation matrix which allows to obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion. That relation was already introduced by Kneller \textit{et al}\cite{Kneller1994} in the context of Lagrangian dynamics for rigid bodies, but here we use it in the derivation of the Hamiltonian equations of motion.

@@

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are the corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. If we differentiate both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}

The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.

\subsection{Euler Parameters and Unit Quaternions}

It is possible to represent the orientation of a rigid body by Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$ that satisfies\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of the Euler parameters is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0.
\end{equation}

This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$. Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}

Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

@@

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}
As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations of motion for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} The equation above is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = -\tr{\mt \Omega}{\vt q}$ and, therefore,
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}

Now this is convenient for differentiation with respect to $\vt q$.  Hence, after carrying out the differentiations as described in Appendix \ref{sec:Diff_Rules} and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following system of differential equations to describe the dynamics of a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = -\diff{U}{\vt r} \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega - \diff{U}{\vt q}
\end{align}
\end{subequations}

For a rigid body composed of individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.

@@

\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{i L_{NVE} \Delta t} = e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L_a \frac{\Delta t}{2}},
\end{equation}

where $L_r = \sum_{k=1}^3 L_k$ is the Liouville operator for a complete rotation and, $i L_a $, $i L_k $ as well as $i L_t $ retain its previous definitions. Recall that each exponential operator in Eq.~\ref{eq:trotter_splitting_NVE} will act having the result of the preceding operator (the one at its right hand side) as initial condition. Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is

\begin{equation}
\label{eq:splitting_rot}
e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n
\end{equation}

where $\delta t = \nicefrac{\Delta t}{n}$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step to improve accuracy. An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, commonly used in Nos\'{e}-Hoover chain dynamics. 
The factorization of the operator is expressed as follows

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
\mathchardef\mhyphen="2D
\usepackage{booktabs}
\usepackage{multirow}
\begin{document}

@@

Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations. This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

@@

one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient. In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble. The property of being symplectic implies the preservation of phase space volume. This feature, together with time reversibility are mandatory for numerical integrators used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}, etc. It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.

In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform MD. The contribution to the NVE case relies on a factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}. In the NVT case, both the translational and rotational degrees of freedom (DOF) are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF. We have also simplified the calculation of the pressure for a rigid body system, whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions (PBC). The expression we derive depends on the resultant interaction forces between rigid body pairs. This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces.

This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy. Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy. Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

\section{Mechanics of Rigid Body Motion}
\label{sec:mechanics}

In this section we first review certain aspects of the kinematics or rigid bodies we consider relevant to this study. In Sec~\ref{sec:eulerparameters}, we introduce one particular factorization of the rotation matrix from which we obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion.

@@

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal inertia axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are the corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. If we differentiate both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}
The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.

%\subsection{Euler Parameters and Unit Quaternions}
\subsection{The Rate of change of Euler Parameters}
\label{sec:eulerparameters}

Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters. One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in terms of those angles. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$. The mathematical relationship between the Euler angles and the quaternion places the restriction over $\boldsymbol{q}$:\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}
Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of the Euler parameters is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0.
\end{equation}
This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}
Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}
This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

@@

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}
In the previous subsection we obtained the relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}), which is used in this part to develop the Hamiltonian formalism of the equations of motion. As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations of motion for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}
Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = -\tr{\mt \Omega}{\vt q}$ and, therefore,
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}
Now this is convenient for differentiation with respect to $\vt q$.  Hence, after carrying out the differentiations as described in Appendix \ref{sec:Diff_Rules} and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following system of differential equations to describe the dynamics of a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = -\diff{U}{\vt r} \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega - \diff{U}{\vt q}
\end{align}
\end{subequations}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.

@@

\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{i L_{NVE} \Delta t} = e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L_a \frac{\Delta t}{2}},
\end{equation}
where $L_r = \sum_{k=1}^3 L_k$ is the Liouville operator for a complete rotation and, $i L_a $, $i L_k $ as well as $i L_t $ retain its previous definitions. Recall that each exponential operator in Eq.~\ref{eq:trotter_splitting_NVE} will act having the result of the preceding operator (the one at its right hand side) as initial condition. Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is\cite{Tuckerman1992}

\begin{equation}
\label{eq:splitting_rot}
e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n,
\end{equation}
where $\delta t = \nicefrac{\Delta t}{n}$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step to improve accuracy. An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, given by
